### 2021-08-21 plos_latex_template.tex
This was mostly reading Janani's, Heather's, and Matt's edits :)

In July 2021, the authors of this article, along with a larger group of people, organized a global and virtual conference for users and developers of the R software, useR! 2021. useR! conferences are strictly non-profit and are organized by volunteers from the R community and supported by the R Foundation. These global meetings include attendees from both academia and the industry. This year, we aimed at building a high-quality conference in a kind, inclusive, accessible, and welcoming environment for everyone. In this article, we share a few critical lessons learned during the past year's work organizing this conference. We highlight these as the 10 simple rules to host an inclusive conference. These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but can also be easily adapted and applied to regional or local conferences.

@@

Conferences are spaces to meet and reconnect with members from a specific community, learn about new advances in the field and share our recent contributions. The larger the conference, the larger the opportunities to network and learn from your cohort. However, just like in science and academia in general, conferences can become an exclusive space for privileged groups (e.g., white, male, from a rich country, English-native speaker, with no physical disabilities)  \cite{arend_disparity_2019, timperley_he_2020, gewin_what_2019, brown_ableism_2018}.

Calls to improve diversity in many academic spaces, including conferences, are becoming frequent as communities and scientific societies are called out for overrepresenting specific groups of people. While these systematic issues are beginning to be addressed, there is scope for much improvement all around.

Diversity encompasses multiple dimensions: age, body ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. There are implicit hierarchies along these axes, and some statuses (e.g., being cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems are consciously and unconsciously built. Thinking about diversity and inclusion is thinking about counteracting the structures that hold these hierarchies in place.

@@

This article suggests rules for pivoting traditional conferences towards inclusiveness and diversity and welcoming minoritized groups. These tips stem from the authors' experience in organizing a global and virtual conference for users and developers of the R programming language, useR! 2021. We aimed at building a high-quality conference not just in terms of technical content but embracing everyone in a kind, inclusive, accessible, and welcoming environment, critical in the context of the COVID-19 pandemic. Most of the authors also have experience organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR conference, summarized as 10 simple rules towards an inclusive conference.

\section*{Rule 1: Set realistic but bold diversity indicators}
.
As a starting point, identify which groups are usually unsupported or discriminated against in the community or event of interest, and determine the contributing factors. Next, set measurable indicators of success when organizing a diverse conference. The numbers resulting from these indicators are not the primary goal, and this is not about performative allyship. Instead, these numbers should guide and help hold the organizing team accountable towards these diverse and inclusive goals.

We found it greatly helpful to publish a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) and express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) and in every decision you make, will let people know that they are seen, respected, and welcomed, and that this is their space and community too. However, it is important to not stop with a statement alone, but identify specific goals and tasks that ensure compliance with these larger goals, for e.g., identifying accessible platforms, developing a detailed code of conduct, and accessibility guidelines (see other Rules below).

\section*{Rule 2: Have an inclusive and diverse organizing team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually underrepresented groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. Even if creating and maintaining such an interdisciplinary team is challenging, the positive outcomes far outweigh any minor inconveniences, and the team will grow, learn, and embrace more inclusive practices. For instance, if the organizing team is global and remote, time zones can be hard to manage, with meetings scheduled at odd hours. The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best platforms to mechanisms to account for every culture and situation. 
A successful well-oiled inclusive team will provide a safe space for everyone not just to learn and grow but identify critical aspects that are working or failing to improve inclusiveness and accessibility together \cite{hong_groups_2004}.

@@

\section*{Rule 3: Finding the best people for each task}

@@

@@

Conferences, even virtual ones, should have registration fees for two primary reasons. The first one is that preparing the conference involves a lot of concerted effort and costs money (e.g., commercial registration tools, captioning, or conference venue, if in-person) and this translates to the participation fee. The second reason has to do with the human psychology -- people value more the things they pay for, and there is a lower attendance rate for free events with registration /cite{eventbrite_ultimate_2017}[any publication to back this up?]. On the other hand, if we are aiming for inclusiveness and representation, the socioeconomic context of participants should be taken into account when determining the registration rates. Usually, there is a higher fee for people from the industry than for academia (~ proportionate to income). We should also consider a lower fee for non-profit organizations, government employees, or freelancers, and use conversion factors (for country of residence using data from International Comparison Program report of the World Bank \cite{arend_disparity_2019}). It is important to include discounts for students as well as for postdocs (early career trainees) to encourage their participation \cite{sarabipour_evaluating_2020, andalib_postdoc_2018, kaplan_postdoc_2012}. Since postdoc (or trainee) statuses are not always well-defined in academia and can vary for each country, their payment category should be explicitly defined. Disabilities may also increase the overall cost of conferences, especially in-person \cite{de_picker_rethinking_2020, irish_increasing_2020}; discounts would help reduce the financial burden for these factions too.

@@

\section*{Rule 5: Do not give scholarships or grants to attend. Give fee waivers or discounts without tedious application forms.}

Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people in need, the language needs to make that clear without shrouding it with fancy-sounding phrases. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. People with limited resources already have a hard time applying for loans, grants, and scholarships all their lives. We should not make the attendees' lives more challenging by asking for long essays to convince us that they deserve our support. \st{Many people that cannot afford international conferences do not even try to apply for waivers/scholarships because they are discouraged by the requirements, the process, and partly because they believe it would be hard for them to win financial/travel aid.} For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. Also, the participant metadata (e.g., demographics, gender, career stage) is already a strong indicator of their potential need for aid.
If the conference resources allow it, the organizers could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

\section*{Acknowledgments}
The authors of this piece would like to thank every single member of the organizing team of useR! 2021 [insert full list of organizers <LINK>] for their valuable contribution to an inclusive conference experience, and the R Foundation for their financial support for useR! 2021.

>>>>

In July 2021, the authors of this article, along with a larger group of people, organized the annual user conference of the R Project for Statistical Computing. This instance of the useR! conference was particularly well received as a virtual and explicitly global event, reaching users and developers of the R language from more than 120 countries. useR! conferences are non-profit events, organized by volunteers from the R community and supported by the R Foundation. The conference attracts a broad range of participants from academia, industry, government and the non-profit sector. For 2021, we aimed to build a high-quality conference in a kind, inclusive, accessible, and welcoming environment for everyone. In this article, we share a few critical lessons learned in the process. We streamline our most important learnings in 10 simple rules to host an inclusive conference. These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but also apply at the regional or local level.

@@

Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field and share our recent contributions. The larger the conference, the larger the opportunities to network and learn from your cohort. However, conferences can become an exclusive space for privileged groups (e.g., white, male, from a rich country, English-native speaker, with no physical disabilities)  \cite{arend_disparity_2019, timperley_he_2020, gewin_what_2019, brown_ableism_2018}.

Calls to improve diversity in many spaces, including conferences, are becoming more frequent as communities and scientific societies are getting called out for over-representing specific groups. While there has been some action to address systematic inequalities, there is still a lot of room for improvement.

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. There are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems are consciously and unconsciously built. Thinking about diversity and inclusion is thinking about counteracting the structures that hold these hierarchies in place.

@@

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. These tips stem from the authors' experience of organizing a virtual and global conference for users and developers of the R programming language, useR! 2021. We embraced the challenge of organizing a high-quality conference in the context of the COVID-19 pandemic using the chance of a virtual conference environment to create a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of  organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

\section*{Rule 1: Set realistic but bold diversity indicators}

As a starting point, identify which groups are usually unsupported or discriminated against in the community or event of interest, and determine the contributing factors. Next, set measurable indicators of success when organizing a diverse conference. The numbers resulting from these indicators are not the primary goal, and this is not about performative allyship. Instead, these numbers should guide and help hold the organizing team accountable towards these diverse and inclusive goals.

Publishing a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) and expressing this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) and in every decision you make, will let people know that they are seen, respected, and welcome; that this is their space and community too. However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct, and preparing accessibility guidelines (see other Rules below).

\section*{Rule 2: Have an inclusive and diverse organizing team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually underrepresented groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. Even if creating and maintaining such an interdisciplinary team is challenging, the positive outcomes far outweigh any minor inconveniences, and the team will grow, learn, and embrace more inclusive practices. For instance, if the organizing team is global and remote, time zones can be hard to manage, with meetings scheduled at odd hours. The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best platforms to mechanisms to account for every culture and situation. 
A successful, inclusive team will not only be a safe space for people to learn and grow but also may help identify critical aspects that are working or failing to improve inclusiveness and accessibility together \cite{hong_groups_2004}.

@@

\section*{Rule 3: Search proactively the best people for each task}

@@

Disabled people often repeat "Nothing about us without us". This means the actual life experiences of people from URM are not replaceable with good intentions from people outside these groups. The same holds for other dimensions of diversity. No space will be truly inclusive and welcoming to URM without people from these groups -and their experiences-. This means people in privileged groups may need to step down from positions of power. Of course, no-one has the same experiences and experiences the same oppressions in life. When building a diverse team, bear in mind that

@@

Conferences, even virtual ones, should have registration fees for two primary reasons. The first one is that preparing the conference involves a lot of concerted effort and costs money (e.g., commercial registration tools, captioning, or conference venue, if in-person) and this translates to the participation fee. The second reason has to do with the human psychology -- people value more the things they pay for, and there is a lower attendance rate for free events with no registration costs \cite{eventbrite_ultimate_2017}. On the other hand, if we are aiming for inclusiveness and representation, the socioeconomic context of participants should be taken into account when determining the registration rates. Usually, there is a higher fee for people from the industry than for academia (~ proportionate to income). We should also consider a lower fee for non-profit organizations, government employees, or freelancers, and use conversion factors (for country of residence using data from International Comparison Program report of the World Bank \cite{arend_disparity_2019}). It is important to include discounts for students as well as for postdocs (early career trainees) to encourage their participation \cite{sarabipour_evaluating_2020, andalib_postdoc_2018, kaplan_postdoc_2012}. Since postdoc (or trainee) statuses are not always well-defined in academia and can vary for each country, their payment category should be explicitly defined. Disabilities may also increase the overall cost of conferences, especially in-person \cite{de_picker_rethinking_2020, irish_increasing_2020}; discounts would help reduce the financial burden for these factions too.

@@

\section*{Rule 5: Abandon diversity scholarships}

%rationale: 1. they might select for the already privileged or people who dont need those to be nice on the cvs 2. they select for outgoing extroverted privileged ("brave") people 3. they are time and emotionally consuming which selects out some people 4. they single out people. even if you controlled for 1 to 3 you would "badge" people - you are creating a minority while trying to avoid it.
Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people who really need it, the language needs to make that clear without shrouding it with fancy-sounding phrases. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. People with limited resources already have a hard time applying for loans, grants, and scholarships. We should not make the attendees' lives more challenging by asking for long essays to convince us that they deserve our support. \st{Many people that cannot afford international conferences do not even try to apply for waivers/scholarships because they are discouraged by the requirements, the process, and partly because they believe it would be hard for them to win financial/travel aid.} For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. Also, the participant metadata (e.g., demographics, gender, career stage) is already a strong indicator of their potential need for aid.
If the conference resources allow it, the organizers could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

\section*{Acknowledgments}
The authors of this piece would like to thank every single member of the organizing team of useR! 2021 [insert full list of organizers \url{https://user2021.r-project.org/about/global-team/}] for their valuable contribution to an inclusive conference experience, and the R Foundation for their financial support for useR! 2021.
